---
title: "Week 1: Cluster Analysis: Exercise 2"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Initialise

# Clear environment
rm(list=ls())
# Clear all plots
if(!is.null(dev.list())) dev.off()

# Install packages
library(biclust)

```


```{r}

dat <- read.table("food.txt", header = TRUE)
head(dat)

# exract relavent variables
dat <- dat[,1:7]

# divide each variable by the weight of the food item
dat<-dat/dat$weight.grams

# remove weight column
dat<-dat[,-6]

# standardize
dat <- scale(dat)
head(dat)

```


```{r}

# Apply Plaid models 
X <- discretize(dat)   # prepping data for plaid model
set.seed(555)
res <- biclust(X, 
               method = BCPlaid(),  # perform Plaid algorithm
               cluster="b",         # cluster both rows and columns
               fit.model=y~m+a+b,   # linear model formula to fit each layer
               verbose = T,
               back.fit = 1,        # to refine fitting of layer via additional iteration runs 
               col.release=0.5,     # threshold to prune columns in the layers depending on column homogeneity
               # higher the threshold, the more are pruned, less clusters selected
               shuffle=5,           # Before a layer is added, it's statistical significance is compared against 5 layers obtained by random 
               max.layers=15)

summary(res)

```
Only one clusters was found. It has 279 observations and only 2 variables. 

```{r}
# Visualise

parallelCoordinates(x=dat, bicResult = res, number = 1, xlim=c(1,290), 
                    ylim=c(-1,5)) # only 1 cluster, so only look at the first cluster
drawHeatmap(x=dat, bicResult = res, number = 1)


```

We can see from the first plot (the line plot) that the bold lines represent the two variables selected for our cluster. The y-axis shows the values obtained for each variable for the different food items selected for the cluster. We can see that they both follow a similar pattern, with food items in the cluster obtaining low values for both of these variables. 

The heatmap again re-iterates all of food items in the cluster obtained low values for both of these variables, illustrated by how the heatmap is mainly made up of light green with very little variation. We can also see that the two variables selected for the cluster were protien and cholestral. 



