---
title: 'Week 1: Multidimensional Scaling Quetsion 3: Non-metric Scaling'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, results='hide'}
# Clear environment
rm(list=ls())
# Clear all plots
if(!is.null(dev.list())) dev.off()

# Install packages
# install.packages('psychTools')
library(psychTools)
# install.packages("smacof")
library(smacof)
# install.packages("Iso")
library(Iso)

# Load data
data(cities)
cities
```

# Question 1:

- Input the dissimilarity ("cities") data from the "psych" library and save it to a variable named "Delta".

```{r, results='markup'}

Delta <- cities 
Delta

```

# Question 2:

- Save the lower triangular matrix of "Delta", as a vector, to a "Diss" variable.

```{r, results='markup'}

Diss <- Delta[lower.tri(Delta)]
Diss

```

# Question 3:

- Label/name the "Diss" variable entries from 1:(M = number of dissimilarities).

```{r, results='markup'}

names(Diss)<-1:length(Diss)  # m=55
Diss

```

# Question 4:

- Order the dissimilarity vector "Diss" in ascending order. Save the corresponding output in a variable denoted "SortedDiss".

```{r, results='markup'}

SortedDiss<-Diss[order(Diss, decreasing = F)]
# or could use sort(Diss, decreasing = FALSE)
SortedDiss

```

# Question 5:

- Use the cmdscale function to obtain an initial configuration of the cities within a 2-dimensional space (k = 2), by means of a classical scaling approach.Save the corresponding output in a variable named Y0. Hint: The function requires a euclidean distance matrix input. 

```{r}

Y0 <- cmdscale(Delta)

```

# Question 6:

- Plot the initial configuration and associated text labels.

```{r, results='markup',fig.pos="H", fig.align='center'}

require(graphics)
Y0_x <- Y0[,1]
Y0_y <- Y0[,2]
plot(Y0_x,Y0_y,
     type="n",
     xlab="First principal co-ordinate",
     ylab="Second principal co-ordinate", 
     asp=1, 
     axes=T, 
     main="cmdscale")
text(Y0_x,Y0_y,rownames(Delta),cex=0.6)

```

# Question 7:

- Calculate the euclidean distances between the initial configuration Y0, and save the output to a variable named "Dist". 

```{r, results='markup'}

Dist <- dist(Y0, method = "euclidean") # just the distances, not the whole matrix
Dist

```

# Question 8:

- Convert the distances to a single vector labelled "VecDist".

```{r, results='markup'}

VecDist <- as.vector(Dist)
VecDist

```

# Question 9:

- Label/name the "VecDist" variable entries from 1:M.

```{r, results='markup'}

names(VecDist)<-1:length(VecDist)

```

# Question 10:

- Order the Y0 distances "VecDist" according to the ordered dissimilarity vector "SortedDiss". Save the output in a variable named "SortedDist". Hint: Use names(SortedDiss) to order the vector

```{r, results='markup'}

SortedDist <- VecDist[names(SortedDiss)]
SortedDist

```

# Question 11:

- Calculate suitable disparities, related to "SortedDist", using isotonic regression. Hint: Apply the pava function, from the "Iso" package, to the "SortedDist" distances. 

```{r}

disp<-pava(SortedDist)

```

# Question 12:

- Plot the "SortedDist" against the index order. Overlay the plot with disparity points, linked by a line. This is known as a Shephard Diagram

```{r, results='markup',fig.pos="H", fig.align='center'}

#plot(x=1:length(SortedDist), y=VecDist, xlab="Dissimilarities", ylab="Distances",
     #main="Shepard Diagram", type="p", pch=16, col="grey")
plot(SortedDist, xlab="Rank Order", ylab="Distances",
     main="Shepard Diagram", type="p", pch=16, col="red")
lines(1:length(SortedDist), disp, pch=16)
lines(1:length(SortedDist), disp, pch=16, type = "b")

```

# Question 13:

- Build Kruskal's stress function with 2 input variables: Y0 distances, disparities. Save the function as "Stress".

```{r, results='markup'}

Stress <- function(Y0_distances,disparities){
  s <- sum((Y0_distances-disparities)^2)
  s <- s/sum(Y0_distances^2)
  s <- sqrt(s)
  s
}

```

# Question 14:

- Calculate the initial stress.

```{r, results='markup'}

Stress(VecDist,disp)

```

# Question 15:

- Redefine the stress function as a function of Y only. Name the function "StressY". Hint: Combine steps 7 - 11 with the stress function.

```{r, results='markup'}


StressY <- function(Y){
  Y0 <- cmdscale(Y)
  
  dist <- dist(Y0, diag = T, upper = T)
  vecdist <- as.matrix(dist)[lower.tri(dist)]
  vecdist <- as.vector(vecdist)
  names(vecdist)<-1:length(vecdist)
  
  diss <- Delta[lower.tri(Y)]
  names(diss)<-1:length(diss)  
  sorteddiss<-diss[order(diss, decreasing = F)]
  
  sorteddist <- vecdist[names(sorteddiss)]
  disp<-pava(sorteddist)
  
  s <- sum((vecdist-disp)^2)
  s <- s/sum(vecdist^2)
  s <- sqrt(s)
  s
}

StressY(Delta)

```

# Question 16:

- Build a plotting function to plot the 2-dimensional co-ordinate output, with text labels. Call this function "PlotCoordinates".

```{r, results='markup',fig.pos="H", fig.align='center'}

PlotCoordinates <- function(Y){
  
  y0 <- cmdscale(Y)
  
  y0_x <- y0[,1]
  y0_y <- y0[,2]
  plot(y0_x,y0_y,type="n",xlab="Dimension x",ylab="Dimension y", asp=1, axes=T, main="cmdscale")
  text(y0_x,y0_y,rownames(Y),cex=0.6)
  
}

PlotCoordinates(Delta)

```

# Question 17:

- Optimise the "StressY" function, return the output using the first 5 optimisation methods present in the "optim" function.

```{r}

OptimizedStressY <- function(Y, initial_configuration){
  # calculate Euclidean distances and vectorise
  Dist_opt <- dist(initial_configuration, method = "euclidean") 
  VecDist_opt <- as.vector(Dist_opt)
  
  # rank order vectorised Euclidean distances
  labels_dist_opt <- 1:length(VecDist_opt)
  names(VecDist_opt) <- labels_dist_opt
  
  # order vectorised Euclidean distances according to ordered dissimilarity vector
  Diss_opt <- as.vector(Y[lower.tri(Y)]) 
  names(Diss_opt) <- 1:length(Diss_opt)
  SortedDiss_opt <- sort(Diss_opt)
  SortedDiss_opt <- VecDist_opt[order(match(names(VecDist_opt), names(SortedDiss_opt)))]
  
  # calculate disparities via isotonic regression
  disparities_opt <- pava(SortedDiss_opt)
  
  # obtain Kruskal's Stress Stress Formula 1 components
  stress <- sqrt((sum((VecDist_opt - disparities_opt)^2))/(sum(VecDist_opt^2)))
  
  return(stress)
}

# initialisation of configuration - classical scaling algorithm 
start_configuration <- cmdscale(cities)
start_configuration <- as.vector(start_configuration)

# Method 1: Nelder-Mead
Y0_optim1 <- optim(par = start_configuration, OptimizedStressY, method = "Nelder-Mead", Y = cities, control = list(maxit = 1000))
Y0_optim1$value
Y0_optim1$convergence

# Method 2: BFGS
Y0_optim2 <- optim(par = start_configuration, OptimizedStressY, method = "BFGS", Y = cities, control = list(maxit = 1000))
Y0_optim2$value
Y0_optim2$convergence

# Method 3: CG
Y0_optim3 <- optim(par = start_configuration, OptimizedStressY, method = "CG", Y = cities, control = list(maxit = 1000))
Y0_optim3$value
Y0_optim3$convergence

# Method 4: L-BFGS-B
Y0_optim4 <- optim(par = start_configuration, OptimizedStressY, method = "L-BFGS-B", Y = cities, control = list(maxit = 1000))
Y0_optim4$value
Y0_optim4$convergence

# Method 5: SANN
Y0_optim5 <- optim(par = start_configuration, OptimizedStressY, method = "SANN", Y = cities, control = list(maxit = 1000))
Y0_optim5$value
Y0_optim5$convergence

```

# Question 18:

- Compare the convergence property and stress values of the prevailing optimisations and plot the lowest stress value configuration.

Note: higher stress AS COMPARED TO the intial configuration DUE TO the ranking of distances/disparities.

```{r}
min(Y0_optim1$value, Y0_optim2$value, Y0_optim3$value, Y0_optim4$value, Y0_optim5$value)

# Belongs to Nelder-Mead

Y_vals <- matrix(Y0_optim1$par, ncol = 2)

plot(Y_vals[,1], Y_vals[,2], type = "p", pch = 20, xlab = "Dimension 1", ylab = "Dimension 2", xlim = c(-2000, 2000), ylim = c(-2000, 2000))
text(Y_vals[,1], Y_vals[,2], labels = names(cities), pos = 4)

```